NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2017-12-01 Manuscript/manuscriptPS.tex
Updates related to the separation of manuscripts

In NMRlipids I and II project we were looking for a MD model
which would correctly reproduce headgroup and glycerol
backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
Here we extend the same goal for other than PC lipids.
Currently the focus is on PE, PG and PS bilayers and their
mixtures with PC. Experimental data with different amounts of 
added salt is now collected and presented in this manuscript.
Simulations for bilayers containing PG and PS with low salt conditions
are ran with parameters from CHARMM GUI
and compared to experiments. Some order parameters from
simulations seems to be off from experimental values.
Simulation data from other models would be highly useful 
to see if some of the existing models would reproduce the 
experimental order parameters and to analyze different 
conformations predicted by different models respect to 
experiments.

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  \caption{\label{lipids}
    Chemical structures and labels for the headgroup carbons.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparameters.eps}
  \caption{\label{HGorderParameters}
    Absolute values of order parameters for headgroup and glycerol backbone with different headgroups
    from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
    POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
    DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    PC headgroup order parameters from experiments of mixtures with
    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
\end{figure}

Absolute values of experimental order parameters for different lipid headgroups are
collected in Fig. \ref{HGorderParameters}. Signs are measured only for PC as far as I know,
thus only absolute values are used for now.

Based on superficial reading, the conclusions in the literature are roughly \\ 
1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
In contrast to PC lipids (see \cite{botan15} and references therein).

@@

\section{Experimental glycerol backbone and headgroup order parameters for PE, PG and PS lipids}

\section{Glycerol backbone and headgroup order parameters for PE, PG and PS lipids in simulations}

Several simulations containing PE, PG and PS lipids have been published \cite{??}, \todo{List should be completed}
however, glycerol backbone and headgroup order parameters are not compared to
the experiments (based on superficial reading of literature).

\subsection{PS headgroup}
The headgroup order parameters of DOPS and POPS bilayers 
from different experiments and simulations are shown in
Figs. \ref{HGorderParametersPS}.
None of the tested models gives satisfactory agreement with
experiments for order parameters in headgroup $\alpha$ and $\beta$
carbons.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.eps}
  \caption{\label{HGorderParametersPS}
    Order parameters for DOPS headgroup and glycerol
    backbone from simulations with different models and experiments without CaCl$_2$ 
    Experimental data from \cite{browning80} contains 0.1M of NaCl.
    Signs are taken from experiments for POPS described in Supplementary Information.
  }
  \todo{Check and report all the counterions.}
  \todo{Glycerol backbone order parameters should be available from the spectra measured by Tiago Ferreira.}
\end{figure}

@@

%Comparison between experiments and simulations with CHARMM GUI
%parameters for PS lipids in Fig. \ref{HGorderParametersPOPS} suggest that
%the model overestimates magnitudes in $\beta$ and g$_3$ order parameters
%and forking in $\alpha$.

\section{Lipid bilayers with mixed compositions}

\subsection{Effect of PE, PS and PG on PC headgroup}
The headgroup order parameters for PC lipids (POPC and DOPC)
mixed with PE, PS and PG lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
from different simulation model and experiments \cite{scherer87} with different
mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
This is not the case in simulation data shown in Fig. \ref{HGorderparametersPCvsPEPSPG}.
The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
smaller than in experiments.

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPEPSPG}
    PC headgroup order parameters from mixtures with PE, PS and PG
    lipids with various mole fractions from different simulation models and experiments \cite{scherer87}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure}

\subsection{Effect of PC on PS and PG headgroups}
The headgroup order parameters for PS and PG lipid mixtures with PC
having different mole fractions from simulations and experiments \cite{borle85,roux90}
are shown in Fig. \ref{HGorderparametersPSPGvsPC}. The effect of increasing
amount of PC to PS headgroup seems to qualitatively incorrect in CHARMM36 simulations.
The $\beta$-carbon order parameter increases in experiment, but decreases
in simulations with both tested counterions (Na+ and K+). Larger
$\alpha$-carbon order parameter decreases with the addition of PC
in experiment, while the lower remains unchanged. In simulations the
larger increases and the lower decreases.
Interestingly, the $\alpha$-carbon order parameters are closer to experiments
in pure PS system with K+ counterions than with Na+. 
The changes in PG headgroup order parameters are minor in simulations, which
is in line with the only available experiment for the $\beta$-carbon.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSPGvsPC}
    PS and PG lipid headgroup order parameters from mixtures with PC
    lipids with various mole fractions from different simulation models and experiments \cite{borle85,roux90}.
    Signs are not yet known experimentally for PG, thus the signs give by simulations are used.
    Signs for PS are measures as described in SI.
  }
  \todo{Some simulations contain potassium as counterions, while some sodium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.} \\
  \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
  pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
\end{figure}

\section{Ca$^{2+}$ binding in bilayers with negatively charged PG and PS lipids}

PC lipid headgroup order parameters can used to measure ion binding
affinity, because their magnitude is linearly proportional
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
This molecular electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Figs \ref{OrderParametersWithCaCl},
\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
as a function of CaCl$_2$ concentration in the presence of different amounts of
negatively charged PS or PG lipids.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
%  \caption{\label{OrderParametersWithCaCl}
%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
%  }
%  \todo{Check the NaCl concentrations in the samples.}
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
%  \caption{\label{OrderParametersWithCaClBELOW1M}
%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
%  }
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
%  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
%    The change of PC headgroup order parameters
%    in the presence of different amount of negatively charged lipids
%    respect to the values without added CaCl$_2$.
%    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
%  }
%\end{figure}

PC headgroup order parameters increase when negatively charged
PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
as expected based on electrometer concept \cite{seelig87}
(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%In electrometer concept this is explained by the tilting of
%headgroup more parallel to membrane normal \cite{??}.
Further, the order parameters decrease with the addition
of CaCl$_2$ and the decrease is more pronounced for systems with more
negatively charged lipids (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
the Ca2+ binding presumably fully cancels the charge from negative lipids and
overcharging occurs above these concenterations.
The interpretation of this data and some other results has been that \cite{seelig90}
\begin{displayquote}
  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
    (ii) the free Ca$^{2+}$
    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
    charge, either due to protein or to lipid; \\
    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
    of bound Ca$^{2+}$ at neutral and charged lipids; \\
    (iv) the actual binding step can be described by a Langmuir
    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
    used to describe the chemical binding equilibrium.''}
\end{displayquote}

\section{Ca$2+$ binding in bilayers with negatively charged PG and PS lipids in simulations}

Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
is shown in Fig. \ref{changesWITHCaClPG}. The decrease of $\alpha$ order parameter
is in agreement with experiments, while decerase of $\beta$ order parameter is overestimated.
The result is very similar to the results with PC in NMRlipids II publication \cite{catte16}.
It should be, however, noted that the $\beta$-order parameters are not actually measured for PG,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
parameter of bound charge \todo{Should we check this against cationic surfactant experiments \cite{scherer89}?}.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaClPGPS.eps}
  \caption{\label{changesWITHCaClPG}
    PG order parameters as a function CaCl$_2$ concentration from experiments \cite{borle85} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured. 
  }
\end{figure}

@@

>>>>

In NMRlipids I and II project we were looking for a MD model
which would correctly reproduce headgroup and glycerol
backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
Here we extend the same goal for lipids with negatively charged PS headgroup.
Chemical structure of PS headgroup together with other common biological
lipids is shown in Fig. \ref{lipids}.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  \caption{\label{lipids}
    Chemical structures and labels for the headgroup carbons.
  }
\end{figure}

Absolute values of experimental order parameters for different lipid headgroups are
collected from the literature in Fig. \ref{HGorderParameters}. Since order parameter
signs are known only for PC, only absolute values are shown.
Main conclusions regarding the structure of different common lipid headgroups in the literature are \\
1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
Careful discussion and analysis of structural details of PE, PG or PS headgroups is not available, 
in contrast to PC lipids (see \cite{botan15} and references therein).
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparameters.eps}
  \caption{\label{HGorderParameters}
    Absolute values of order parameters for headgroup and glycerol backbone with different headgroups
    from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
    POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
    DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
  }
\end{figure}

As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
surfactants. The changes with the addition of neutral lipids is significantly smaller.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    PC headgroup order parameters from experiments of mixtures with
    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
\end{figure}

@@

\section{Results and Discussion}

\subsection{PS headgroup and glycerol backbone structure in simulations and experiments}
The headgroup order parameters of DOPS and POPS bilayers 
from different experiments and simulations are shown in
Figs. \ref{HGorderParametersPS}.
None of the tested models gives satisfactory agreement with
experiments for order parameters in headgroup $\alpha$ and $\beta$
carbons.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.eps}
  \caption{\label{HGorderParametersPS}
    Order parameters for DOPS headgroup and glycerol
    backbone from simulations with different models and experiments without CaCl$_2$ 
    Experimental data from \cite{browning80} contains 0.1M of NaCl.
    Signs are taken from experiments for POPS described in Supplementary Information.
  }
  \todo{Check and report all the counterions.}
  \todo{Glycerol backbone order parameters should be available from the spectra measured by Tiago Ferreira.}
\end{figure}

@@

\todo{Dihedral angle distribtions in Fig. \ref{dihedrals} should be included in the discussion.}

\subsection{Effect of PS on PC headgroup}
The headgroup order parameters of POPC 
mixed with PS lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
from different simulation model and experiments \cite{scherer87} with different
mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
This is not the case in simulation data shown in Fig. \ref{HGorderparametersPCvsPEPSPG}.
%The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
%order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
%in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
%headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
%smaller than in experiments.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPEPSPG}
    PC headgroup order parameters from mixtures with PE, PS and PG
    lipids with various mole fractions from different simulation models and experiments \cite{scherer87}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure}

\todo{More data to be collected before discussion.}

\subsection{Effect of PC on PS headgroup}
The headgroup order parameters of PS mixed with varying amounts of PC
from simulations and experiments \cite{borle85,roux90}
are shown in Fig. \ref{HGorderparametersPSPGvsPC}. The effect of increasing
amount of PC to PS headgroup seems to qualitatively incorrect in CHARMM36 simulations.
The $\beta$-carbon order parameter increases in experiment, but decreases
in simulations with both tested counterions (Na+ and K+). Larger
$\alpha$-carbon order parameter decreases with the addition of PC
in experiment, while the lower remains unchanged. In simulations the
larger increases and the lower decreases.
Interestingly, the $\alpha$-carbon order parameters are closer to experiments
in pure PS system with K+ counterions than with Na+. 
%The changes in PG headgroup order parameters are minor in simulations, which
%is in line with the only available experiment for the $\beta$-carbon.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSPGvsPC}
    PS order parameters from mixtures with PC
    lipids with various mole fractions from different simulation models and experiments \cite{borle85,roux90}.
    Signs for PS are measures as described in SI.
  }
  \todo{Some simulations contain potassium as counterions, while some sodium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.} \\
  \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
  pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
\end{figure}

\subsection{Ca$^{2+}$ binding affinity in bilayers with negatively charged PS lipids}

PC lipid headgroup order parameters can used to measure ion binding
affinity, because their magnitude is proportional
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
The molecular electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
%\ref{OrderParametersWithCaClBELOW1M} and
\ref{OrderParameterCHANGESWithCaClBELOW1M},
showing the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
as a function of CaCl$_2$ concentration in the presence of different amounts of
negatively charged PS or PG lipids.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
%  \caption{\label{OrderParametersWithCaCl}
%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
%  }
%  \todo{Check the NaCl concentrations in the samples.}
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
%  \caption{\label{OrderParametersWithCaClBELOW1M}
%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
%  }
%\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    The change of PC headgroup order parameters
    in the presence of different amount of negatively charged lipids
    respect to the values without added CaCl$_2$.
    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
  }
\end{figure}
%PC headgroup order parameters increase when negatively charged
%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
%as expected based on electrometer concept \cite{seelig87}
%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%In electrometer concept this is explained by the tilting of
%headgroup more parallel to membrane normal \cite{??}.
The decrease of order parameters with CaCl$_2$
is more pronounced for systems with more
negatively charged lipids.% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
overcharging occurs above these concenterations.
The interpretation of this data and some other results has been that \cite{seelig90}
\begin{displayquote}
  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
    (ii) the free Ca$^{2+}$
    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
    charge, either due to protein or to lipid; \\
    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
    of bound Ca$^{2+}$ at neutral and charged lipids; \\
    (iv) the actual binding step can be described by a Langmuir
    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
    used to describe the chemical binding equilibrium.''}
\end{displayquote}

\todo{When we have more data for Ca binding to PS containing bilayers, the discussion will be updated and PG results moved
to other manuscript.}
Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
is shown in Fig. \ref{changesWITHCaClPG}. The decrease of $\alpha$ order parameter
is in agreement with experiments, while decerase of $\beta$ order parameter is overestimated.
The result is very similar to the results with PC in NMRlipids II publication \cite{catte16}.
It should be, however, noted that the $\beta$-order parameters are not actually measured for PG,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
parameter of bound charge
\todo{The response of CHARMM36 against cationic surfactant experiments \cite{scherer89} is to be checked.
  I have already ran the simulations, analysis yet to be done}.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaClPGPS.eps}
  \caption{\label{changesWITHCaClPG}
    PG order parameters as a function CaCl$_2$ concentration from experiments \cite{borle85} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured. 
  }
\end{figure}

@@

\subsection{Effect of Ca$2+$ binding to PS headgroup}
